---
title: "Querying and Summarizing Census PUMS Data in R"
author: "Will Carter"
---

# Introduction
We will manipulate and process data sets from Public Use Microdata Sample (PUMS) census API. We are focusing on datasets for years 2021, 2022, 2023, and 2024.


```{r}
#| label: setup

# tidyverse attaches dplyr, tibble, purrr, ggplot2, and lubridate. httr makes the web requests and jsonlite parses JSON.
library(httr)
library(jsonlite)
library(tidyverse)
```

# Part 1: Data Processing

The API returns mapped values. A person's `SEX` returns `1` or `2`, etc. The context behind these mappings are published in the `variables.json` file from the census bureau. Each variable in `variables.json` has a `label` and a `values` entry. Categorical variables' `values` holds `item` list of code and the description, whereas continuous variables hold a `range` of min, max, and description. 

The first helper lists the varaibles needed and defines `sort_by_code()` because the JSON files stores codes in not in order, but the factor levels should follow the order of the codes.

```{r}
#| label: variable-helpers

# variables the project needs. The state variable is named ST in the 2021 and
# 2022 files and STATE in the 2023 and 2024 files, so it is added separately
# inside parse_pums_variables().
pums_vars_needed <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP",
                      "PWGTP", "FER", "HHL", "SCH", "SCHL", "SEX",
                      "REGION", "DIVISION")

# order a named vector of code -> description by numeric code. If any code
# isn't numeric, leave the original order alone.
sort_by_code <- function(items) {
  codes <- suppressWarnings(as.numeric(names(items)))
  if (is.null(items) || anyNA(codes)) return(items)
  items[order(codes)]
}
```

The parsing function takes the JSON after it has been read into an R list and
returns one element per needed variable. Each element is itself a list with
the variable's `label`, its `items` , and its `ranges`. The state
variable is stored under the single name `STATE`, so downstream ingestors do not need to account for which year the data is from.

```{r}
#| label: parse-variables

# pull the needed variables out of one parsed variables.json.
# parsed: list from fromJSON(..., simplifyVector = FALSE)
# year: single survey year (decides whether to look for ST or STATE)
parse_pums_variables <- function(parsed, year) {
  all_vars  <- parsed$variables
  state_var <- if (year <= 2022) "ST" else "STATE"
  wanted    <- c(pums_vars_needed, state_var)

  missing_vars <- setdiff(wanted, names(all_vars))
  if (length(missing_vars) > 0) {
    stop("Variables not found in the ", year, " file: ",
         paste(missing_vars, collapse = ", "))
  }

  out <- lapply(all_vars[wanted], function(v) {
    list(
      label  = v$label,
      items  = sort_by_code(unlist(v$values$item)),
      ranges = lapply(v$values$range, function(r) {
        list(min = as.numeric(r$min), max = as.numeric(r$max),
             description = r$description)
      })
    )
  })

  # store the state variable under one name for every year
  names(out)[names(out) == state_var] <- "STATE"
  out
}
```

The second function checks the year, builds the URL for
that year's `variables.json`, requests it with `GET()`, stops on an HTTP
error, and passes the parsed JSON to the function above. Keeping the request
and the parsing in separate functions means the parser can be tested on a
saved file without touching the API.

```{r}
#| label: get-variables

# GET the variables.json for one year and parse it.
get_pums_variables <- function(year) {
  if (length(year) != 1 || !(year %in% 2021:2024)) {
    stop("year must be a single value: 2021, 2022, 2023, or 2024")
  }
  url <- paste0("https://api.census.gov/data/", year,
                "/acs/acs1/pums/variables.json")
  resp <- GET(url)
  stop_for_status(resp)
  parsed <- fromJSON(content(resp, as = "text", encoding = "UTF-8"),
                     simplifyVector = FALSE)
  parse_pums_variables(parsed, year)
}
```

## Checking function on year 2024

Check the parser output. Input year 2024. The output should be 15 variables and `SEX` should show its two codes `Male` and `Female`.

```{r}
#| label: example-one-year

vars_2024 <- get_pums_variables(2024)
names(vars_2024)
vars_2024$SEX$items
```

The names are the 14 variables the project asks for plus `STATE`, and `SEX`
has two codes: `1` is Male and `2` is Female.

The state variable is called `ST` in the 2021 and 2022 files. The parser
should hide that difference, so asking for 2021 should still give a list that
contains `STATE`.

```{r}
#| label: example-older-year

vars_2021 <- get_pums_variables(2021)
"STATE" %in% names(vars_2021)
```

The year check should reject anything outside 2021 to 2024. The
chunk below is set to keep rendering after an error so the message shows on
the page.

```{r}
#| label: example-bad-year
#| error: true

get_pums_variables(2020)
```

The parser was shown to run successfully for 2024. Now we run it on all years
and save the results for later code to reference.


## Building the lists for all four years

One list per year is built and saved to `data/pums_variables.rds`. The
`if` statement prevents redownloading on rerenders.The saved lists are
then read back in as `pums_vars`.

```{r}
#| label: build-variable-lists

if (!file.exists("data/pums_variables.rds")) {
  pums_variable_lists <- lapply(2021:2024, get_pums_variables)
  names(pums_variable_lists) <- 2021:2024
  dir.create("data", showWarnings = FALSE)
  saveRDS(pums_variable_lists, "data/pums_variables.rds")
}

pums_vars <- readRDS("data/pums_variables.rds")
```

Check the lists. The list is indexed by year and each year holds the 15 variables.

```{r}
#| label: check-variable-lists

names(pums_vars)
names(pums_vars[["2024"]])
pums_vars[["2024"]]$SEX$items
length(pums_vars[["2021"]]$STATE$items)
```

The first two lines confirm the four years and the 15 variables per year. The
third shows the mappings for `SEX`. The last shows that the state list has 52 entries, the 50 states, DC, and Puerto Rico. Puerto Rico will be dropped when state input is validated.



## Parsing an API response
 
A request to the PUMS API consists of the base address for the year, a
`get=` list of variables, a `for=` clause for the geography, any optional
filters, and the key. Before writing the main query function, one small
request shows what the API sends back. This one asks for sex, person weight,
age, arrival time at work, travel time to work, and education for people in
North Carolina (`for=state:37`) who are 30 years old with a 45 minute commute
(`AGEP=30&JWMNP=45`), which keeps the response short.
 
The key is read from the `CENSUS_KEY` environment variable instead of being
typed into the document, because this code is published with the site.
 
```{r}
#| label: example-request
 
census_key <- Sys.getenv("CENSUS_KEY")
stopifnot("Set CENSUS_KEY in .Renviron and restart R" = nzchar(census_key))
 
url <- paste0("https://api.census.gov/data/2024/acs/acs1/pums",
              "?get=SEX,PWGTP,AGEP,JWAP,JWMNP,SCHL",
              "&for=state:37&AGEP=30&JWMNP=45",
              "&key=", census_key)
resp <- GET(url)
status_code(resp)
cat(substr(content(resp, as = "text", encoding = "UTF-8"), 1, 300))
```
 
A status of `200` means the request was successful. In the JSON body returned, the first array is the header and each following array is one person's record. Every value arrives as a string. `AGEP` and `JWMNP` appear twice. The API repeats any variable used as a filter at the end of each row. The geography comes back as a column named `state`.
 
The first helper turns the JSON text into a tibble. It uses the first row as
the column names and drops any repeated column. First it checks if the return is JSON to ensure there is no error response.
 
```{r}
#| label: parse-census-json
 
# turn the text of a Census API reply into a tibble of character columns.
# the result is an array of arrays: row 1 is the header, the rest is data.
parse_census_json <- function(json_text) {
  if (!validate(json_text)) {
    stop("The Census API did not return JSON. The reply began: ",
         substr(json_text, 1, 200), call. = FALSE)
  }
  mat <- fromJSON(json_text)             # a character matrix
  colnames(mat) <- mat[1, ]
  # Variables used as filters are repeated at the end; keep the first copy
  mat <- mat[-1, !duplicated(colnames(mat)), drop = FALSE]
  as_tibble(mat)
}
```
 
The second helper works on the object `GET()` returns. It checks the HTTP
status and stops with the API's own message when something is wrong. For
example, a misspelled variable name produces a status `400` with a short
explanation, and a query that matches no records produces a `204`. Otherwise
it passes the text to the first helper.
 
```{r}
#| label: response-to-tibble
 
# convert the object returned by GET() into a tibble
pums_response_to_tibble <- function(resp) {
  reply <- content(resp, as = "text", encoding = "UTF-8")
  if (status_code(resp) == 204) {
    stop("The query matched no records.", call. = FALSE)
  }
  if (status_code(resp) != 200) {
    stop("Census API request failed (status ", status_code(resp), "): ",
         reply, call. = FALSE)
  }
  parse_census_json(reply)
}
```
 
Applying it to the response from above gives a tibble with one row per
person.
 
```{r}
#| label: example-tibble
 
census_raw <- pums_response_to_tibble(resp)
census_raw
```
 
The repeated `AGEP` and `JWMNP` columns are dropped, `state` is kept, and every
column is consistently a character vector.






## Converting variable types
 
The tibble from the last section is all text. Numeric variables need to be
numbers and categorical variables need to be factors with labels. Both helpers use the lists saved earlier in `pums_vars`, so the labels always match
the survey year.

Some numeric codes mean "not applicable". For `JWMNP` (travel time to work), `0` means the person is not a worker or worked from home, and in 2023 `888` means the value was suppressed. `GASP` and `GRPIP` have similar codes. If they remained as numbers, they would skew every mean, so they become `NA`. The helper finds them from the labels (a label starting with "N/A" or "Suppressed"). A real zero is kept: for `AGEP`, `0` means under 1 year old. Codes have to be compared as numbers. The API sends `1` where the variables file has `01`, and the 2021 file uses `00` for an age of 0. Matching as text would miss these, so both sides are converted to numbers first.
 
For categorical variables the N/A code is kept as its own level.
 
```{r}
#| label: type-helpers
 
# codes that mean "not applicable" or "suppressed" for one variable. they are
# found from the labels, so the rule works for every year's file
na_codes_for <- function(var_info) {
  items <- var_info$items
  if (length(items) == 0) return(character(0))
  names(items)[grepl("^(N/A|Suppressed)", items)]
}
 
# character codes -> numbers, with the variable's N/A codes set to NA
convert_numeric <- function(x, var_info) {
  values <- as.numeric(x)
  values[values %in% as.numeric(na_codes_for(var_info))] <- NA
  values
}
 
# character codes -> factor whose levels are the variable's labels, in code
# order. codes are matched as numbers. unknown codes trigger a warning
convert_factor <- function(x, var_info) {
  codes   <- as.numeric(x)
  items   <- var_info$items
  known   <- as.numeric(names(items))
  unknown <- setdiff(unique(codes[!is.na(codes)]), known)
  if (length(unknown) > 0) {
    warning("Codes not found in the variable list: ",
            paste(unknown, collapse = ", "), call. = FALSE)
  }
  factor(codes, levels = known, labels = unname(items))
}
```
 
A few direct calls show the behavior. The first turns the N/A and suppressed
codes into `NA`, the second keeps a real zero, and the third shows that the
API's `1` finds the file's `01`.
 
```{r}
#| label: example-conversions
 
convert_numeric(c("12", "0", "888"), pums_vars[["2023"]]$JWMNP)
convert_numeric(c("0", "5"), pums_vars[["2024"]]$AGEP)
as.character(convert_factor(c("1", "16", "22"), pums_vars[["2024"]]$SCHL))
```
 
Applying the helpers to the tibble converts the weight, age, and travel time to numbers and `SEX` and `SCHL` to factors.
 
```{r}
#| label: example-typed
 
lists_2024 <- pums_vars[["2024"]]
 
census_typed <- census_raw |>
  mutate(
    PWGTP = convert_numeric(PWGTP, lists_2024$PWGTP),
    AGEP  = convert_numeric(AGEP,  lists_2024$AGEP),
    JWMNP = convert_numeric(JWMNP, lists_2024$JWMNP),
    SEX   = convert_factor(SEX,  lists_2024$SEX),
    SCHL  = convert_factor(SCHL, lists_2024$SCHL)
  )
 
census_typed
levels(census_typed$SEX)
```
 
`PWGTP`, `AGEP`, and `JWMNP` are now numeric (`<dbl>`), and `SEX` is a factor
with the levels Male and Female. `JWAP`. The arrival time at work is still
text and will use a helper to convert to datetime.


## Time variables
 
We have two times variables. `JWAP` (when the person arrives at work) and `JWDP` (when they leave for work). Neither comes as a clock time. Each value is a code for a window of time, such as "7:30 a.m. to 7:34 a.m.". To use them as times, each window is replaced by its midpoint.
 
`JWAP` uses 5 minute windows, while `JWDP` uses 5, 10, and 30 minute windows. So the helper reads the start and end times out of each label. The midpoint is the time halfway between the two times in the label, so the midpoint of "7:30 a.m. to 7:34 a.m." is 7:32. Codes that mean "not applicable" (not a worker, or worked from home) or "suppressed" have no window, so they become `NA`.
 
The result is a `lubridate` period, which stores the time of day as hours, minutes, and seconds after midnight.
 
```{r}
#| label: time-helpers
 
# "7:30 a.m." -> seconds after midnight. 12 a.m. is hour 0 and 12 p.m. is
# hour 12, which is what the %% 12 handles.
clock_to_seconds <- function(clock) {
  hour   <- as.numeric(sub(":.*", "", clock))
  minute <- as.numeric(sub("^\\d{1,2}:(\\d{2}).*", "\\1", clock))
  pm     <- grepl("p\\.m\\.", clock)
  (hour %% 12 + ifelse(pm, 12, 0)) * 3600 + minute * 60
}
 
# "7:30 a.m. to 7:34 a.m." -> midpoint in seconds after midnight, or NA when
# the label has no window (N/A and suppressed codes).
label_to_midpoint <- function(label) {
  clocks <- regmatches(label, gregexpr("\\d{1,2}:\\d{2} [ap]\\.m\\.", label))[[1]]
  if (length(clocks) != 2) return(NA_real_)
  start <- clock_to_seconds(clocks[1])
  end   <- clock_to_seconds(clocks[2])
  if (end < start) end <- end + 86400     # a window that crosses midnight
  (start + (end - start) / 2) %% 86400
}
 
# character codes -> lubridate periods holding each window's midpoint.
# codes are matched as numbers, as in the other helpers.
convert_time <- function(x, var_info) {
  items     <- var_info$items
  midpoints <- unname(vapply(items, label_to_midpoint, numeric(1)))
  midpoints[names(items) %in% na_codes_for(var_info)] <- NA
  seconds_to_period(midpoints[match(as.numeric(x), as.numeric(names(items)))])
}
```
 
In the 2024 file the code `088` is the window 7:30 a.m. to 7:34 a.m., and the API sends it as `88`.
 
```{r}
#| label: example-time
 
pums_vars[["2024"]]$JWAP$items[["088"]]
convert_time("88", pums_vars[["2024"]]$JWAP)
convert_time("0", pums_vars[["2024"]]$JWAP)
```
 
Applying it to the tibble converts the arrival time.
 
```{r}
#| label: example-timed
 
census_timed <- census_typed |>
  mutate(JWAP = convert_time(JWAP, lists_2024$JWAP))
 
census_timed
```
 
`JWAP` is now a period column. The first row's code `88` became 7:32, the
middle of its window, and a person with no arrival time would show `NA`. Every variable type the project needs can be converted with our helper functions.

 
 
## Querying one year
 
The helpers so far can: turn a response into a tibble, and
convert codes into numbers, factors, and times. The single-year query function
chains them. It checks the user's inputs, builds the request URL from the survey year, the variables,
and the geography, sends the request with `GET()`, turns the reply into a
tibble, and converts each column to its proper type using that year's variable
lists.
 
The first small helper builds the URL.
 
```{r}
#| label: build-url
 
# Build the request URL. The key is part of it, so don't print the result.
build_pums_url <- function(year, vars, for_clause, key) {
  paste0("https://api.census.gov/data/", year, "/acs/acs1/pums",
         "?get=", paste(vars, collapse = ","),
         "&for=", for_clause,
         "&key=", key)
}
```
 
The second helper applies the right conversion to each requested column.
 
```{r}
#| label: convert-types
 
# convert each requested column using one year's variable lists.
convert_pums_types <- function(data, year_lists, numeric_vars, cat_vars) {
  time_vars <- c("JWAP", "JWDP")
 
  for (var in numeric_vars) {
    data[[var]] <- if (var %in% time_vars) {
      convert_time(data[[var]], year_lists[[var]])
    } else {
      convert_numeric(data[[var]], year_lists[[var]])
    }
  }
  for (var in cat_vars) {
    data[[var]] <- convert_factor(data[[var]], year_lists[[var]])
  }
  data
}
```
 
### Checking the inputs
 
The query should fail early, with a message that says what to fix, instead of sending a bad request to the API. `year` must be a single value from 2021 to 2024. `numeric_vars` may only contain `AGEP`, `GASP`, `GRPIP`, `JWAP`, `JWDP`, `JWMNP`, and `PWGTP`, and needs at least one variable other than `PWGTP`. The person weight `PWGTP` is always returned, so it is added if the user leaves it out. `cat_vars` may only contain `FER`, `HHL`, `SCH`, `SCHL`, and `SEX`, and needs at least one.
 
The first helper checks one set of variable names against its allowed values, so the same code serves both arguments.
 
```{r}
#| label: check-choices
 
pums_numeric_options <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP")
pums_cat_options     <- c("FER", "HHL", "SCH", "SCHL", "SEX")
 
# stop unless `values` is a non-empty character vector made only of `allowed`
# names. the message names the offending values and lists the options.
check_choices <- function(values, allowed, arg_name) {
  if (length(values) == 0) {
    stop(arg_name, " must include at least one variable. Choose from: ",
         paste(allowed, collapse = ", "), ".", call. = FALSE)
  }
  if (!is.character(values) || anyNA(values)) {
    stop(arg_name, " must be a character vector of variable names.",
         call. = FALSE)
  }
  bad <- setdiff(values, allowed)
  if (length(bad) > 0) {
    stop(arg_name, " cannot include ", paste(bad, collapse = ", "),
         ". Choose from: ", paste(allowed, collapse = ", "), ".",
         call. = FALSE)
  }
}
```
 
The second helper runs all the checks and returns the cleaned arguments, with `PWGTP` added to the numeric variables if it was missing.
 
```{r}
#| label: validate-args
 
# check the user's inputs and return them cleaned up.
validate_pums_args <- function(year, numeric_vars, cat_vars) {
  if (length(year) != 1 || is.na(year) || !(year %in% 2021:2024)) {
    stop("year must be a single value: 2021, 2022, 2023, or 2024",
         call. = FALSE)
  }
  check_choices(numeric_vars, c(pums_numeric_options, "PWGTP"), "numeric_vars")
  check_choices(cat_vars, pums_cat_options, "cat_vars")
 
  numeric_vars <- unique(c(numeric_vars, "PWGTP"))   # PWGTP is always returned
  if (length(setdiff(numeric_vars, "PWGTP")) == 0) {
    stop("numeric_vars must include at least one variable besides PWGTP. ",
         "Choose from: ", paste(pums_numeric_options, collapse = ", "), ".",
         call. = FALSE)
  }
  list(year = year, numeric_vars = numeric_vars, cat_vars = unique(cat_vars))
}
```
 
### The query function
 
The query function defaults ask for the 2024 survey, `AGEP` and `PWGTP` as the numeric variables, `SEX` as the categorical variable, and North Carolina. The key comes from the `CENSUS_KEY` environment variable, and the variable lists default to the `pums_vars` object loaded earlier. The inputs are validated first so that a bad request fails before call to API.
 
```{r}
#| label: get-pums-year
 
# query the PUMS API for one year and return a tibble with converted types.
get_pums_year <- function(year = 2024,
                          numeric_vars = c("AGEP", "PWGTP"),
                          cat_vars = "SEX",
                          for_clause = "state:37",
                          key = Sys.getenv("CENSUS_KEY"),
                          var_lists = pums_vars) {
  args <- validate_pums_args(year, numeric_vars, cat_vars)
  url  <- build_pums_url(args$year, c(args$numeric_vars, args$cat_vars),
                         for_clause, key)
  data <- pums_response_to_tibble(GET(url))
  convert_pums_types(data, var_lists[[as.character(args$year)]],
                     args$numeric_vars, args$cat_vars)
}
```
 
Calling it with no arguments returns every North Carolina record in the 2024
survey.
 
```{r}
#| label: example-default-query
 
nc_2024 <- get_pums_year()
nc_2024
```
 
The result has one row per sampled person, with `AGEP` and `PWGTP` as numbers,
`SEX` as a factor, and the geography column `state` from the API. In our second call it asks for Wyoming (`"state:56"`), in 2022, with the weight, travel time, and arrival time as the numeric variables and sex and education as the categorical variables.
 
```{r}
#| label: example-custom-query
 
wyoming_2022 <- get_pums_year(
  year = 2022,
  numeric_vars = c("PWGTP", "JWMNP", "JWAP"),
  cat_vars = c("SEX", "SCHL"),
  for_clause = "state:56"
)
wyoming_2022
```
 
The columns are returned in the order requested and each have the correct type.
 
If the user leaves out the weight, the function adds it. Here only `JWMNP` is requested and `PWGTP` still comes back.
 
```{r}
#| label: example-weight-added
 
names(get_pums_year(numeric_vars = "JWMNP", for_clause = "state:56"))
```
 
### Calls that should fail
 
Each call below breaks one rule, and the validation checks are run before anything is sent to the API. Set to keep going after an error so that every message shows.
 
```{r}
#| label: example-errors
#| error: true
 
get_pums_year(year = 2020)                              # year out of range
get_pums_year(year = c(2022, 2023))                     # more than one year
get_pums_year(numeric_vars = "PWGTP")                   # nothing besides PWGTP
get_pums_year(numeric_vars = c("AGEP", "SEX"))          # SEX is categorical
get_pums_year(numeric_vars = "WAGP")                    # not an allowed variable
get_pums_year(cat_vars = character(0))                  # no categorical variable
get_pums_year(cat_vars = c("SEX", "AGEP"))              # AGEP is numeric
```
 
Each message names what was wrong and lists the allowed choices.