---
title: "Querying and Summarizing Census PUMS Data in R"
author: "Will Carter"
---

# Introduction
We will manipulate and process data sets from Public Use Microdata Sample (PUMS) census API. We are focusing on datasets for years 2021, 2022, 2023, and 2024.


```{r}
#| label: setup

# tidyverse attaches dplyr, tibble, purrr, ggplot2, and lubridate. httr makes the web requests and jsonlite parses JSON.
library(httr)
library(jsonlite)
library(tidyverse)
```

# Part 1: Data Processing

The API returns mapped values. A person's `SEX` returns `1` or `2`, etc. The context behind these mappings are published in the `variables.json` file from the census bureau. Each variable in `variables.json` has a `label` and a `values` entry. Categorical variables' `values` holds `item` list of code and the description, whereas continuous variables hold a `range` of min, max, and description. 

The first helper lists the varaibles needed and defines `sort_by_code()` because the JSON files stores codes in not in order, but the factor levels should follow the order of the codes.

```{r}
#| label: variable-helpers

# variables the project needs. The state variable is named ST in the 2021 and
# 2022 files and STATE in the 2023 and 2024 files, so it is added separately
# inside parse_pums_variables().
pums_vars_needed <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP",
                      "PWGTP", "FER", "HHL", "SCH", "SCHL", "SEX",
                      "REGION", "DIVISION")

# order a named vector of code -> description by numeric code. If any code
# isn't numeric, leave the original order alone.
sort_by_code <- function(items) {
  codes <- suppressWarnings(as.numeric(names(items)))
  if (is.null(items) || anyNA(codes)) return(items)
  items[order(codes)]
}
```

The parsing function takes the JSON after it has been read into an R list and
returns one element per needed variable. Each element is itself a list with
the variable's `label`, its `items` , and its `ranges`. The state
variable is stored under the single name `STATE`, so downstream ingestors do not need to account for which year the data is from.

```{r}
#| label: parse-variables

# pull the needed variables out of one parsed variables.json.
# parsed: list from fromJSON(..., simplifyVector = FALSE)
# year: single survey year (decides whether to look for ST or STATE)
parse_pums_variables <- function(parsed, year) {
  all_vars  <- parsed$variables
  state_var <- if (year <= 2022) "ST" else "STATE"
  wanted    <- c(pums_vars_needed, state_var)

  missing_vars <- setdiff(wanted, names(all_vars))
  if (length(missing_vars) > 0) {
    stop("Variables not found in the ", year, " file: ",
         paste(missing_vars, collapse = ", "))
  }

  out <- lapply(all_vars[wanted], function(v) {
    list(
      label  = v$label,
      items  = sort_by_code(unlist(v$values$item)),
      ranges = lapply(v$values$range, function(r) {
        list(min = as.numeric(r$min), max = as.numeric(r$max),
             description = r$description)
      })
    )
  })

  # store the state variable under one name for every year
  names(out)[names(out) == state_var] <- "STATE"
  out
}
```

The second function checks the year, builds the URL for
that year's `variables.json`, requests it with `GET()`, stops on an HTTP
error, and passes the parsed JSON to the function above. Keeping the request
and the parsing in separate functions means the parser can be tested on a
saved file without touching the API.

```{r}
#| label: get-variables

# GET the variables.json for one year and parse it.
get_pums_variables <- function(year) {
  if (length(year) != 1 || !(year %in% 2021:2024)) {
    stop("year must be a single value: 2021, 2022, 2023, or 2024")
  }
  url <- paste0("https://api.census.gov/data/", year,
                "/acs/acs1/pums/variables.json")
  resp <- GET(url)
  stop_for_status(resp)
  parsed <- fromJSON(content(resp, as = "text", encoding = "UTF-8"),
                     simplifyVector = FALSE)
  parse_pums_variables(parsed, year)
}
```

## Checking function on year 2024

Check the parser output. Input year 2024. The output should be 15 variables and `SEX` should show its two codes `Male` and `Female`.

```{r}
#| label: example-one-year

vars_2024 <- get_pums_variables(2024)
names(vars_2024)
vars_2024$SEX$items
```

The names are the 14 variables the project asks for plus `STATE`, and `SEX`
has two codes: `1` is Male and `2` is Female.

The state variable is called `ST` in the 2021 and 2022 files. The parser
should hide that difference, so asking for 2021 should still give a list that
contains `STATE`.

```{r}
#| label: example-older-year

vars_2021 <- get_pums_variables(2021)
"STATE" %in% names(vars_2021)
```

The year check should reject anything outside 2021 to 2024. The
chunk below is set to keep rendering after an error so the message shows on
the page.

```{r}
#| label: example-bad-year
#| error: true

get_pums_variables(2020)
```

The parser was shown to run successfully for 2024. Now we run it on all years
and save the results for later code to reference.

