---
title: "Querying and Summarizing Census PUMS Data in R"
author: "Will Carter"
---

# Introduction

The Public Use Microdata Sample (PUMS) data is a set of person-level records from the U.S. Census Bureau served through an API. This report builds a small toolkit in R for working with that data. We focus on the survey years 2021, 2022, 2023, and 2024.

Each row of PUMS data is a sampled person and `PWGTP`, is the person weight used to estimate how many people in the population would be represented by this record. Every summary and plot use `PWGTP`.

The report consists of three parts.

1. **Data processing.** Functions that read the variable descriptions the API publishes, query the API, and return tibbles with numbers, factors, and times.
2. **Generic functions.** `summary()` and `plot()` methods for those tibbles that use the weights.
3. **Investigation.** The functions are used to ask whether commute time and age differ by education level in North Carolina, and whether that changed from 2021 to 2024.

This implementation uses a Census API key stored in the `CENSUS_KEY` environment variable. The `quantreg` package has to be installed for the weighted boxplots.


```{r}
#| label: setup

# tidyverse attaches dplyr, tibble, purrr, ggplot2, and lubridate. httr makes the web requests and jsonlite parses JSON.
library(httr)
library(jsonlite)
library(tidyverse)
```

# Part 1: Data Processing

The API returns mapped values. A person's `SEX` returns `1` or `2`, etc. The context behind these mappings are published in the `variables.json` file from the census bureau. Each variable in `variables.json` has a `label` and a `values` entry. Categorical variables' `values` holds `item` list of code and the description, whereas continuous variables hold a `range` of min, max, and description. 

This helper function lists the varaibles needed and defines `sort_by_code()` because the JSON files do not store the codes in numeric order, while factor levels should follow the code order.

```{r}
#| label: variable-helpers

# variables the project needs. The state variable is named ST in the 2021 and
# 2022 files and STATE in the 2023 and 2024 files, so it is added separately
# inside parse_pums_variables().
pums_vars_needed <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP",
                      "PWGTP", "FER", "HHL", "SCH", "SCHL", "SEX",
                      "REGION", "DIVISION")

# order a named vector of code -> description by numeric code. If any code
# isn't numeric, leave the original order alone.
sort_by_code <- function(items) {
  codes <- suppressWarnings(as.numeric(names(items)))
  if (is.null(items) || anyNA(codes)) return(items)
  items[order(codes)]
}
```

The parsing function takes the JSON after it has been read into an R list and
returns one element per needed variable. Each element is itself a list with
the variable's `label`, its `items` , and its `ranges`. The state
variable is stored under the single name `STATE`, so downstream ingestors do not need to account for which year the data is from.

```{r}
#| label: parse-variables

# pull the needed variables out of one parsed variables.json.
# parsed: list from fromJSON(..., simplifyVector = FALSE)
# year: single survey year (decides whether to look for ST or STATE)
parse_pums_variables <- function(parsed, year) {
  all_vars  <- parsed$variables
  state_var <- if (year <= 2022) "ST" else "STATE"
  wanted    <- c(pums_vars_needed, state_var)

  missing_vars <- setdiff(wanted, names(all_vars))
  if (length(missing_vars) > 0) {
    stop("Variables not found in the ", year, " file: ",
         paste(missing_vars, collapse = ", "))
  }

  out <- lapply(all_vars[wanted], function(v) {
    list(
      label  = v$label,
      items  = sort_by_code(unlist(v$values$item)),
      ranges = lapply(v$values$range, function(r) {
        list(min = as.numeric(r$min), max = as.numeric(r$max),
             description = r$description)
      })
    )
  })

  # store the state variable under one name for every year
  names(out)[names(out) == state_var] <- "STATE"
  out
}
```

The second function checks the year, builds the URL for
that year's `variables.json`, requests it with `GET()`, stops on an HTTP
error, and passes the parsed JSON to the function above. Keeping the request
and the parsing in separate functions means the parser can be tested on a
saved file without touching the API.

```{r}
#| label: get-variables

# GET the variables.json for one year and parse it.
get_pums_variables <- function(year) {
  if (length(year) != 1 || !(year %in% 2021:2024)) {
    stop("year must be a single value: 2021, 2022, 2023, or 2024")
  }
  url <- paste0("https://api.census.gov/data/", year,
                "/acs/acs1/pums/variables.json")
  resp <- GET(url)
  stop_for_status(resp)
  parsed <- fromJSON(content(resp, as = "text", encoding = "UTF-8"),
                     simplifyVector = FALSE)
  parse_pums_variables(parsed, year)
}
```

## Checking function on year 2024

Check the parser output. Input year 2024. The output should be 15 variables and `SEX` should show its two codes `Male` and `Female`.

```{r}
#| label: example-one-year

vars_2024 <- get_pums_variables(2024)
names(vars_2024)
vars_2024$SEX$items
```

The names are the 14 variables as well as `STATE`. `SEX`has two codes: `1` is Male and `2` is Female.

The state variable is called `ST` in the 2021 and 2022 files. The parser should hide that difference, so asking for 2021 should still give a list that contains `STATE`.

```{r}
#| label: example-older-year

vars_2021 <- get_pums_variables(2021)
"STATE" %in% names(vars_2021)
```

The year check should reject anything outside 2021 to 2024. The
chunk below is set to keep rendering after an error so the message shows on
the page.

```{r}
#| label: example-bad-year
#| error: true

get_pums_variables(2020)
```

The parser was shown to run successfully for 2024. Now we run it on all years
and save the results for later code to reference.


## Building the lists for all four years

One list per year is built and saved to `data/pums_variables.rds`. The
`if` statement prevents redownloading on rerenders.The saved lists are
then read back in as `pums_vars`.

```{r}
#| label: build-variable-lists

if (!file.exists("data/pums_variables.rds")) {
  pums_variable_lists <- lapply(2021:2024, get_pums_variables)
  names(pums_variable_lists) <- 2021:2024
  dir.create("data", showWarnings = FALSE)
  saveRDS(pums_variable_lists, "data/pums_variables.rds")
}

pums_vars <- readRDS("data/pums_variables.rds")
```

Check the lists. The list is indexed by year and each year holds the 15 variables.

```{r}
#| label: check-variable-lists

names(pums_vars)
names(pums_vars[["2024"]])
pums_vars[["2024"]]$SEX$items
length(pums_vars[["2021"]]$STATE$items)
```

The first two lines confirm the four years and the 15 variables per year. The
third shows the mappings for `SEX`. The last shows that the state list has 52 entries, the 50 states, DC, and Puerto Rico. Puerto Rico will be dropped when state input is validated.



## Parsing an API response
 
A request to the PUMS API consists of the base address for the year, a `get=` list of variables, a `for=` clause for the geography, any optional filters, and the key. Before writing the main query function, one small request shows what the API sends back. Sex, person weight, age, arrival time at work, travel time to work, and education for people in North Carolina (`for=state:37`) who are 30 years old with a 45 minute commute (`AGEP=30&JWMNP=45`), which keeps the response short.
 
The key is read from the `CENSUS_KEY` environment variable instead of being typed into the document, because this code is published with the site.
 
```{r}
#| label: example-request
 
census_key <- Sys.getenv("CENSUS_KEY")
stopifnot("Set CENSUS_KEY in .Renviron and restart R" = nzchar(census_key))
 
url <- paste0("https://api.census.gov/data/2024/acs/acs1/pums",
              "?get=SEX,PWGTP,AGEP,JWAP,JWMNP,SCHL",
              "&for=state:37&AGEP=30&JWMNP=45",
              "&key=", census_key)
resp <- GET(url)
status_code(resp)
cat(substr(content(resp, as = "text", encoding = "UTF-8"), 1, 300))
```
 
A status of `200` means the request was successful. In the JSON body returned, the first array is the header and each following array is one person's record. Every value arrives as a string. `AGEP` and `JWMNP` appear twice. The API repeats any variable used as a filter at the end of each row. The geography comes back as a column named `state`.
 
The helper turns the JSON text into a tibble. It uses the first row as the column names and drops any repeated column. First it checks if the return is JSON to ensure there is no error response.
 
```{r}
#| label: parse-census-json
 
# turn the text of a Census API reply into a tibble of character columns.
# the result is an array of arrays: row 1 is the header, the rest is data.
parse_census_json <- function(json_text) {
  if (!validate(json_text)) {
    stop("The Census API did not return JSON. The reply began: ",
         substr(json_text, 1, 200), call. = FALSE)
  }
  mat <- fromJSON(json_text)             # a character matrix
  colnames(mat) <- mat[1, ]
  # Variables used as filters are repeated at the end; keep the first copy
  mat <- mat[-1, !duplicated(colnames(mat)), drop = FALSE]
  as_tibble(mat)
}
```
 
This helper works on the object `GET()` returns. It checks the HTTP status and stops with the API's own message when something is wrong. For example, a misspelled variable name produces a status `400` with a short explanation, and a query that matches no records produces a `204`. Otherwise it passes the text to the first helper.
 
```{r}
#| label: response-to-tibble
 
# convert the object returned by GET() into a tibble
pums_response_to_tibble <- function(resp) {
  reply <- content(resp, as = "text", encoding = "UTF-8")
  if (status_code(resp) == 204) {
    stop("The query matched no records.", call. = FALSE)
  }
  if (status_code(resp) != 200) {
    stop("Census API request failed (status ", status_code(resp), "): ",
         reply, call. = FALSE)
  }
  parse_census_json(reply)
}
```
 
Applying it to the response from above gives a tibble with one row per person.
 
```{r}
#| label: example-tibble
 
census_raw <- pums_response_to_tibble(resp)
census_raw
```
 
The repeated `AGEP` and `JWMNP` columns are dropped, `state` is kept, and every column is consistently a character vector.






## Converting variable types
 
The tibble from the last section is all text. Numeric variables need to be numbers and categorical variables need to be factors with labels. Both helpers use the lists saved earlier in `pums_vars`, so the labels always match the survey year.

Some numeric codes mean "not applicable". For `JWMNP` (travel time to work), `0` means the person is not a worker or worked from home, and in 2023 `888` means the value was suppressed. `GASP` and `GRPIP` have similar codes. If they remained as numbers, they would skew every mean, so they become `NA`. The helper finds them from the labels (a label starting with "N/A" or "Suppressed"). A real zero is kept: for `AGEP`, `0` means under 1 year old. Codes have to be compared as numbers. The API sends `1` where the variables file has `01`, and the 2021 file uses `00` for an age of 0. Matching as text would miss these, so both sides are converted to numbers first.

For categorical variables the N/A code is kept as its own level.
 
```{r}
#| label: type-helpers
 
# codes that mean "not applicable" or "suppressed" for one variable. they are
# found from the labels, so the rule works for every year's file
na_codes_for <- function(var_info) {
  items <- var_info$items
  if (length(items) == 0) return(character(0))
  names(items)[grepl("^(N/A|Suppressed)", items)]
}
 
# character codes -> numbers, with the variable's N/A codes set to NA
convert_numeric <- function(x, var_info) {
  values <- as.numeric(x)
  values[values %in% as.numeric(na_codes_for(var_info))] <- NA
  values
}
 
# character codes -> factor whose levels are the variable's labels, in code
# order. codes are matched as numbers. unknown codes trigger a warning
convert_factor <- function(x, var_info) {
  codes   <- as.numeric(x)
  items   <- var_info$items
  known   <- as.numeric(names(items))
  unknown <- setdiff(unique(codes[!is.na(codes)]), known)
  if (length(unknown) > 0) {
    warning("Codes not found in the variable list: ",
            paste(unknown, collapse = ", "), call. = FALSE)
  }
  factor(codes, levels = known, labels = unname(items))
}
```
 
A few direct calls show the behavior. The first turns the N/A and suppressed codes into `NA`, the second keeps a real zero, and the third shows that the API's `1` finds the file's `01`.
 
```{r}
#| label: example-conversions
 
convert_numeric(c("12", "0", "888"), pums_vars[["2023"]]$JWMNP)
convert_numeric(c("0", "5"), pums_vars[["2024"]]$AGEP)
as.character(convert_factor(c("1", "16", "22"), pums_vars[["2024"]]$SCHL))
```
 
Applying the helpers to the tibble converts the weight, age, and travel time to numbers and `SEX` and `SCHL` to factors.
 
```{r}
#| label: example-typed
 
lists_2024 <- pums_vars[["2024"]]
 
census_typed <- census_raw |>
  mutate(
    PWGTP = convert_numeric(PWGTP, lists_2024$PWGTP),
    AGEP  = convert_numeric(AGEP,  lists_2024$AGEP),
    JWMNP = convert_numeric(JWMNP, lists_2024$JWMNP),
    SEX   = convert_factor(SEX,  lists_2024$SEX),
    SCHL  = convert_factor(SCHL, lists_2024$SCHL)
  )
 
census_typed
levels(census_typed$SEX)
```
 
`PWGTP`, `AGEP`, and `JWMNP` are now numeric (`<dbl>`), and `SEX` is a factor with the levels Male and Female. `JWAP`. The arrival time at work is still text and will use a helper to convert to datetime.


## Time variables
 
We have two times variables. `JWAP` (when the person arrives at work) and `JWDP` (when they leave for work). Neither comes as a clock time. Each value is a code for a window of time, such as "7:30 a.m. to 7:34 a.m.". To use them as times, each window is replaced by its midpoint.
 
`JWAP` uses 5 minute windows, while `JWDP` uses 5, 10, and 30 minute windows. The helper reads the start and end times out of each label. The midpoint is the time halfway between the two times in the label, so the midpoint of "7:30 a.m. to 7:34 a.m." is 7:32. Codes that mean "not applicable" (not a worker, or worked from home) or "suppressed" have no window, so they become `NA`.
 
The result is a `lubridate` period, which stores the time of day as hours, minutes, and seconds after midnight.
 
```{r}
#| label: time-helpers
 
# "7:30 a.m." -> seconds after midnight. 12 a.m. is hour 0 and 12 p.m. is
# hour 12, which is what the %% 12 handles.
clock_to_seconds <- function(clock) {
  hour   <- as.numeric(sub(":.*", "", clock))
  minute <- as.numeric(sub("^\\d{1,2}:(\\d{2}).*", "\\1", clock))
  pm     <- grepl("p\\.m\\.", clock)
  (hour %% 12 + ifelse(pm, 12, 0)) * 3600 + minute * 60
}
 
# "7:30 a.m. to 7:34 a.m." -> midpoint in seconds after midnight, or NA when
# the label has no window (N/A and suppressed codes).
label_to_midpoint <- function(label) {
  clocks <- regmatches(label, gregexpr("\\d{1,2}:\\d{2} [ap]\\.m\\.", label))[[1]]
  if (length(clocks) != 2) return(NA_real_)
  start <- clock_to_seconds(clocks[1])
  end   <- clock_to_seconds(clocks[2])
  if (end < start) end <- end + 86400     # a window that crosses midnight
  (start + (end - start) / 2) %% 86400
}
 
# character codes -> lubridate periods holding each window's midpoint.
# codes are matched as numbers, as in the other helpers.
convert_time <- function(x, var_info) {
  items     <- var_info$items
  midpoints <- unname(vapply(items, label_to_midpoint, numeric(1)))
  midpoints[names(items) %in% na_codes_for(var_info)] <- NA
  seconds_to_period(midpoints[match(as.numeric(x), as.numeric(names(items)))])
}
```
 
In the 2024 file the code `088` is the window 7:30 a.m. to 7:34 a.m., and the API sends it as `88`.
 
```{r}
#| label: example-time
 
pums_vars[["2024"]]$JWAP$items[["088"]]
convert_time("88", pums_vars[["2024"]]$JWAP)
convert_time("0", pums_vars[["2024"]]$JWAP)
```
 
Applying it to the tibble converts the arrival time.
 
```{r}
#| label: example-timed
 
census_timed <- census_typed |>
  mutate(JWAP = convert_time(JWAP, lists_2024$JWAP))
 
census_timed
```
 
`JWAP` is now a period column. The first row's code `88` became 7:32, the middle of its window, and a person with no arrival time would show `NA`. Every variable type the project needs can be converted with the helper functions.

 
 
## Querying one year
 
The helper functions can turn a response into a tibble, and convert codes into numbers, factors, and times. The single-year query function chains them. It checks the user's inputs, builds the request URL from the survey year, the variables, and the geography, sends the request with `GET()`, turns the reply into a tibble, and converts each column to its proper type using that year's variable lists.
 
Builds the URL.
 
```{r}
#| label: build-url
 
# Build the request URL. The key is part of it, so don't print the result.
build_pums_url <- function(year, vars, for_clause, key) {
  paste0("https://api.census.gov/data/", year, "/acs/acs1/pums",
         "?get=", paste(vars, collapse = ","),
         "&for=", for_clause,
         "&key=", key)
}
```
 
Apply the right conversion to each requested column.
 
```{r}
#| label: convert-types
 
# convert each requested column using one year's variable lists.
convert_pums_types <- function(data, year_lists, numeric_vars, cat_vars) {
  time_vars <- c("JWAP", "JWDP")
 
  for (var in numeric_vars) {
    data[[var]] <- if (var %in% time_vars) {
      convert_time(data[[var]], year_lists[[var]])
    } else {
      convert_numeric(data[[var]], year_lists[[var]])
    }
  }
  for (var in cat_vars) {
    data[[var]] <- convert_factor(data[[var]], year_lists[[var]])
  }
  data
}
```
 
### Checking the inputs
 
The query should fail early, with a message that says what to fix, instead of sending a bad request to the API. `year` must be a single value from 2021 to 2024. `numeric_vars` may only contain `AGEP`, `GASP`, `GRPIP`, `JWAP`, `JWDP`, `JWMNP`, and `PWGTP`, and needs at least one variable other than `PWGTP`. The person weight `PWGTP` is always returned, so it is added if the user leaves it out. `cat_vars` may only contain `FER`, `HHL`, `SCH`, `SCHL`, and `SEX`, and needs at least one.
 
This checks one set of variable names against its allowed values, so the same code serves both arguments.
 
```{r}
#| label: check-choices
 
pums_numeric_options <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP")
pums_cat_options     <- c("FER", "HHL", "SCH", "SCHL", "SEX")
 
# stop unless `values` is a non-empty character vector made only of `allowed`
# names. the message names the offending values and lists the options.
check_choices <- function(values, allowed, arg_name) {
  if (length(values) == 0) {
    stop(arg_name, " must include at least one variable. Choose from: ",
         paste(allowed, collapse = ", "), ".", call. = FALSE)
  }
  if (!is.character(values) || anyNA(values)) {
    stop(arg_name, " must be a character vector of variable names.",
         call. = FALSE)
  }
  bad <- setdiff(values, allowed)
  if (length(bad) > 0) {
    stop(arg_name, " cannot include ", paste(bad, collapse = ", "),
         ". Choose from: ", paste(allowed, collapse = ", "), ".",
         call. = FALSE)
  }
}
```
 
Runs all the checks and return the cleaned arguments, with `PWGTP` added to the numeric variables if it was missing.
 
```{r}
#| label: validate-args
 
# check the user's inputs and return them cleaned up.
validate_pums_args <- function(year, numeric_vars, cat_vars) {
  if (length(year) != 1 || is.na(year) || !(year %in% 2021:2024)) {
    stop("year must be a single value: 2021, 2022, 2023, or 2024",
         call. = FALSE)
  }
  check_choices(numeric_vars, c(pums_numeric_options, "PWGTP"), "numeric_vars")
  check_choices(cat_vars, pums_cat_options, "cat_vars")
 
  numeric_vars <- unique(c(numeric_vars, "PWGTP"))   # PWGTP is always returned
  if (length(setdiff(numeric_vars, "PWGTP")) == 0) {
    stop("numeric_vars must include at least one variable besides PWGTP. ",
         "Choose from: ", paste(pums_numeric_options, collapse = ", "), ".",
         call. = FALSE)
  }
  list(year = year, numeric_vars = numeric_vars, cat_vars = unique(cat_vars))
}
```


### Choosing the geography
 
The user picks a geography level, `"Region"`, `"Division"`, or `"State"`, and then one place at that level, such as `"North Carolina"`. The API wants this as a `for=` clause such as `state:37`. The helper functions below processes the user's choice. The level has to be exactly one of the three. `"All"` is not a level because of the size of the data. The place is given as a code or as the label stored in the saved lists, such as `37` or `"North Carolina/NC"`, in any capitalization, and the allowed places come from the lists saved in `pums_vars`. The value `"all"` returns every region, division, or state. If no place is given, each level has a default that contains North Carolina.
 
Set the defaults and look up the allowed places for a level.
 
```{r}
#| label: geography-lookup
 
# the place used when the user gives a level but no value
default_geography <- c(region = "South",
                       division = "South Atlantic (South region)",
                       state = "North Carolina/NC")
 
# code -> label for one level, leaving out Puerto Rico
geography_choices <- function(level, year_lists) {
  items <- year_lists[[toupper(level)]]$items
  items[!grepl("^Puerto Rico", items)]
}
```
 
Check the level, find the code for the place the user typed, and put the `for=` clause together. A value that matches nothing stops with a message that shows what is allowed.
 
```{r}
#| label: build-for-clause
 
# check the geography level and return it in lowercase, as the API spells it
check_geography_level <- function(geography) {
  if (length(geography) != 1 || is.na(geography) ||
      !(tolower(geography) %in% c("region", "division", "state"))) {
    stop("geography must be a single value: \"Region\", \"Division\", or ",
         "\"State\". \"All\" is not available because of the size of the data.",
         call. = FALSE)
  }
  tolower(geography)
}
 
# find the code for what the user typed: a code (37) or a label as stored in
# the saved lists (North Carolina/NC), in any capitalization
match_geography <- function(value, items, level) {
  typed <- tolower(trimws(as.character(value)))
  if (grepl("^[0-9]+$", typed)) {
    hit <- as.numeric(names(items)) == as.numeric(typed)
  } else {
    hit <- tolower(items) == typed
  }
  if (!any(hit)) {
    shown <- if (length(items) <= 10) items else head(items, 5)
    stop("\"", value, "\" is not a valid ", level, ". Use a code or a label, ",
         "such as: ", paste(shown, collapse = "; "),
         if (length(items) > 10) "; ..." else "", ".", call. = FALSE)
  }
  names(items)[hit][1]
}
 
# build the for= clause, for example "state:37". "all" becomes a * wildcard.
build_for_clause <- function(level, value, year_lists) {
  if (is.null(value)) value <- default_geography[[level]]
  if (length(value) != 1 || is.na(value)) {
    stop("geography_value must be a single value, or \"all\".", call. = FALSE)
  }
  if (tolower(trimws(as.character(value))) == "all") {
    return(paste0(level, ":*"))
  }
  items <- geography_choices(level, year_lists)
  paste0(level, ":", match_geography(value, items, level))
}
```
 
The API adds a column named `state`, `region`, or `division` to the response. The last helper turns it into a factor of readable labels, in the same way as the other categorical variables. Levels that do not appear are dropped.
 
```{r}
#| label: convert-geography
 
# turn the geography column into a factor of labels
convert_geography <- function(data, level, year_lists) {
  data[[level]] <- convert_factor(data[[level]], year_lists[[toupper(level)]])
  data <- data[!grepl("^Puerto Rico", as.character(data[[level]])), ]
  data[[level]] <- droplevels(data[[level]])
  data
}
```
 
Calling `build_for_clause()` directly shows what the user's input turns into. The same state works as its label or its code, and `"all"` becomes `*`. Running a query with `"all"` downloads the whole country.
 
```{r}
#| label: example-clauses
 
build_for_clause("state", "North Carolina/NC", lists_2024)
build_for_clause("state", "wyoming/wy", lists_2024)
build_for_clause("state", 37, lists_2024)
build_for_clause("region", "South", lists_2024)
build_for_clause("division", "Mountain (West region)", lists_2024)
build_for_clause("state", "all", lists_2024)
build_for_clause("state", NULL, lists_2024)
```
 
### The query function
 
The query function defaults ask for the 2024 survey, `AGEP` and `PWGTP` as the numeric variables, `SEX` as the categorical variable, and the state of North Carolina. The key comes from the `CENSUS_KEY` environment variable, and the variable lists default to the `pums_vars` object loaded earlier. The inputs are validated first so that a bad request fails before call to API. The geography is checked next, before the tibble is returned it is given the class `census`, which is what lets the summary and plot methods in Part 2 recognize it.
 
```{r}
#| label: get-pums-year
 
# query the PUMS API for one year and return a tibble with converted types.
get_pums_year <- function(year = 2024,
                          numeric_vars = c("AGEP", "PWGTP"),
                          cat_vars = "SEX",
                          geography = "State",
                          geography_value = NULL,
                          key = Sys.getenv("CENSUS_KEY"),
                          var_lists = pums_vars) {
  args       <- validate_pums_args(year, numeric_vars, cat_vars)
  year_lists <- var_lists[[as.character(args$year)]]
  level      <- check_geography_level(geography)
  for_clause <- build_for_clause(level, geography_value, year_lists)
 
  url  <- build_pums_url(args$year, c(args$numeric_vars, args$cat_vars),
                         for_clause, key)
  data <- pums_response_to_tibble(GET(url))
  data <- convert_pums_types(data, year_lists, args$numeric_vars, args$cat_vars)
  data <- convert_geography(data, level, year_lists)
 
  class(data) <- c("census", class(data))
  data
}
```
 
Calling it with no arguments returns every North Carolina record in the 2024 survey.
 
```{r}
#| label: example-default-query
 
nc_2024 <- get_pums_year()
nc_2024
```
 
The result has one row per sampled person, with `AGEP` and `PWGTP` as numbers, `SEX` as a factor, and the geography column `state` as a factor of state names. In our second call it asks for Wyoming, in 2022, with the weight, travel time, and arrival time as the numeric variables and sex and education as the categorical variables.
 
```{r}
#| label: example-custom-query
 
wyoming_2022 <- get_pums_year(
  year = 2022,
  numeric_vars = c("PWGTP", "JWMNP", "JWAP"),
  cat_vars = c("SEX", "SCHL"),
  geography_value = "Wyoming/WY"
)
wyoming_2022
```
 
The columns are returned in the order requested and each have the correct type.
 
If the user leaves out the weight, the function adds it. Here only `JWMNP` is requested and `PWGTP` still comes back.
 
```{r}
#| label: example-weight-added
 
names(get_pums_year(numeric_vars = "JWMNP", geography_value = "Wyoming/WY"))
```
 
### Other geography levels
 
The geography level changes with `geography`. This call is for the New England division in 2023, the smallest of the divisions to download. The `division` column holds the division's label, and the result carries the `census` class.
 
```{r}
#| label: example-division-query
 
new_england_2023 <- get_pums_year(year = 2023,
                                  geography = "Division",
                                  geography_value = "New England (Northeast region)")
new_england_2023
inherits(new_england_2023, "census")
```
 
### Calls that should fail
 
The following examples test the main validation errors.
 
```{r}
#| label: example-errors
#| error: true
 
get_pums_year(year = 2020)  # year out of range
get_pums_year(year = c(2022, 2023))  # more than one year
get_pums_year(numeric_vars = "PWGTP")  # nothing besides PWGTP
get_pums_year(numeric_vars = c("AGEP", "SEX")) # SEX is categorical
get_pums_year(numeric_vars = "WAGP")  # not an allowed variable
get_pums_year(cat_vars = character(0))  # no categorical variable
get_pums_year(cat_vars = c("SEX", "AGEP")) # AGEP is numeric
get_pums_year(geography = "All")  # All is not a level
get_pums_year(geography = c("State", "Region"))  # more than one level
get_pums_year(geography = "County")  # not an available level
get_pums_year(geography_value = "Atlantis")  # no such state
get_pums_year(geography = "Region", geography_value = "NC")  # NC is not a region
get_pums_year(geography_value = c("NC", "SC"))  # more than one place
get_pums_year(geography_value = "Puerto Rico")  # not covered by the project
```
 
Each message names what was wrong and lists the allowed choices.


## Querying several years
 
The single-year function does the work for one survey year. The multi-year function calls it once for each requested year, adds a `year` column that records which survey each row came from, and stacks the results into one tibble. The variables, the geography, and the key are passed through to the single-year function with `...`, so those options and their defaults are defined in one place.
 
The years are checked first before anything is downloaded. The `year` column is a factor. Stacking tibbles with `bind_rows()` can drop the custom `census` class, so the function makes sure it is still present before returning.
 
```{r}
#| label: get-pums-years
 
# check the years and return them sorted and without repeats
check_years <- function(years) {
  if (length(years) == 0 || anyNA(years) || !all(years %in% 2021:2024)) {
    stop("years must be one or more of: 2021, 2022, 2023, 2024", call. = FALSE)
  }
  sort(unique(as.integer(years)))
}
 
# query several survey years and stack them into one census tibble.
# ... holds the other options and is passed on to get_pums_year().
get_pums_years <- function(years = 2024, ...) {
  years   <- check_years(years)
  results <- lapply(years, function(y) get_pums_year(year = y, ...))
  names(results) <- years
 
  data      <- bind_rows(results, .id = "year")
  data$year <- factor(data$year, levels = years)
 
  # bind_rows() can drop the custom class, so make sure it is still there
  if (!inherits(data, "census")) class(data) <- c("census", class(data))
  data
}
```
 
This call is for all four years for Wyoming, with the weight and travel time as the numeric variables and sex and education as the categorical variables.
 
```{r}
#| label: example-multi-year
 
wyoming_all_years <- get_pums_years(
  years = 2021:2024,
  numeric_vars = c("PWGTP", "JWMNP"),
  cat_vars = c("SEX", "SCHL"),
  geography_value = "Wyoming/WY"
)
wyoming_all_years
table(wyoming_all_years$year)
inherits(wyoming_all_years, "census")
```
 
The table shows how many records each survey year contributed, and the object still has the `census` class. Years can be given in any order and with repeats. They come back sorted, one copy each.
 
```{r}
#| label: example-year-order
 
levels(get_pums_years(years = c(2024, 2022, 2024),
                      geography_value = "Wyoming/WY")$year)
```
 
### Calls that should fail
 
The year check runs before any download, and the other options are validated by the single-year function before it makes its first request.
 
```{r}
#| label: example-year-errors
#| error: true
 
get_pums_years(years = 2020)  # year out of range
get_pums_years(years = c(2023, 2025))  # one good year and one bad year
get_pums_years(years = integer(0))  # no years
get_pums_years(years = c(2022, 2023), numeric_vars = "PWGTP")  # nothing besides PWGTP
```




# Part 2: Generic Functions
 
The records in our tibbles do not correspond to just one person. The `PWGTP` variable is the number of people a row represents, so every summary and plot has to weight by it. Repeating each row `PWGTP` times would give an ordinary data set, but it would make some tibbles huge, so the functions below use the weights directly.
 
A generic function such as `summary()` or `plot()` looks at the class of its first argument and calls the method written for that class. The query functions return tibbles with the class `census`, so a function named `summary.census()` is what `summary()` runs on them.
 
```{r}
#| label: example-class
 
class(nc_2024)
```
 
## Summarizing a census tibble
 
The summary method reports a mean and a standard deviation for each numeric variable and counts for each categorical variable. By default it uses every numeric variable except `PWGTP` and every factor in the tibble, including `year` and the geography column. The user can instead name the variables to summarize. The time variables are left out.
 
The mean and standard deviation use the weighted formulas from the assignment document, where $x_i$ is a value and $w_i$ is its weight (`PWGTP`):
 
$$\bar{x} = \frac{\sum x_i w_i}{\sum w_i}, \qquad s = \sqrt{\frac{\sum x_i^2 w_i}{\sum w_i} - \bar{x}^2}$$
 
Some people do not have a value for a variable, such as travel time for someone who does not work. Those rows are left out of that variable's mean and standard deviation along with their weights, so the result describe people who have a valid value. Counts are also weighted. The count for a level is the sum of the weights in that level, which estimates the number of people. The number of sample records at each level is returned as well.
 
Do the arithmetic.
 
```{r}
#| label: weighted-helpers
 
# weighted mean of x with weights w. missing values are left out with their weights.
weighted_mean <- function(x, w) {
  keep <- !is.na(x)
  sum(x[keep] * w[keep]) / sum(w[keep])
}
 
# weighted standard deviation from the project's formula. rounding can push
# the variance slightly below zero when every value is the same, so it is
# held at zero.
weighted_sd <- function(x, w) {
  keep     <- !is.na(x)
  variance <- sum(x[keep]^2 * w[keep]) / sum(w[keep]) - weighted_mean(x, w)^2
  sqrt(max(variance, 0))
}
 
# estimated number of people at each level of a factor: the weights added up
# within each level. levels with no records count as 0.
weighted_counts <- function(x, w) {
  totals <- tapply(w, x, sum)
  totals[is.na(totals)] <- 0
  c(totals)
}
```
 
The summary method decides which variables can be summarized, checks the user's choices with the same `check_choices()` used earlier, and returns a named list. 
 
```{r}
#| label: summary-census
 
# summary method for census tibbles. returns a named list:
#   means, sds - weighted mean and standard deviation of each numeric variable
#   counts     - estimated number of people at each level of each factor
#   records    - number of sample records at each level of each factor
summary.census <- function(object, numeric_vars = NULL, cat_vars = NULL) {
  if (!"PWGTP" %in% names(object)) {
    stop("the data needs a PWGTP column to use as weights.", call. = FALSE)
  }
 
  # what can be summarized: numeric columns other than the weight (times are
  # left out) and factor columns
  is_number <- vapply(object, function(col) is.numeric(col) && !is.period(col),
                      logical(1))
  is_factor <- vapply(object, is.factor, logical(1))
  numeric_options <- setdiff(names(object)[is_number], "PWGTP")
  cat_options <- names(object)[is_factor]
 
  if (is.null(numeric_vars)) {
    numeric_vars <- numeric_options
  } else {
    check_choices(numeric_vars, numeric_options, "numeric_vars")
  }
  if (is.null(cat_vars)) {
    cat_vars <- cat_options
  } else {
    check_choices(cat_vars, cat_options, "cat_vars")
  }
 
  weights <- object$PWGTP
  list(
    means = vapply(numeric_vars, function(v) weighted_mean(object[[v]], weights),
                   numeric(1)),
    sds = vapply(numeric_vars, function(v) weighted_sd(object[[v]], weights),
                   numeric(1)),
    counts = sapply(cat_vars, function(v) weighted_counts(object[[v]], weights),
                     simplify = FALSE),
    records = sapply(cat_vars, function(v) c(table(object[[v]])),
                     simplify = FALSE)
  )
}
```
 
Calling `summary()` on the North Carolina data with no other arguments summarizes `AGEP` and the two factors, `SEX` and the geography column.
 
```{r}
#| label: example-summary-default
 
summary(nc_2024)
```
 
The multi-year Wyoming data has travel time as its numeric variable, and `year` is one of the factors, so the counts are split by survey year.
 
```{r}
#| label: example-summary-years
 
summary(wyoming_all_years)
```
 
The user can pick the variables. This call summarizes only travel time and counts only `year` and `SEX`.
 
```{r}
#| label: example-summary-chosen
 
summary(wyoming_all_years, numeric_vars = "JWMNP", cat_vars = c("year", "SEX"))
```
 
The earlier Wyoming 2022 data also has arrival time, `JWAP`. It is a time, so the default summary skips it and reports only travel time.
 
```{r}
#| label: example-summary-time
 
summary(wyoming_2022)$means
```
 
The weights change the answer. The first number below treats every record as one person. The second counts each record as the `PWGTP` people it stands for.
 
```{r}
#| label: example-weighted-vs-plain
 
mean(wyoming_all_years$JWMNP, na.rm = TRUE)
summary(wyoming_all_years)$means["JWMNP"]
```
 
### Calls that should fail
 
The following examples test the main validation errors.
 
```{r}
#| label: example-summary-errors
#| error: true
 
summary(wyoming_all_years, numeric_vars = "SEX")  # SEX is a factor
summary(wyoming_all_years, numeric_vars = "PWGTP")  # the weight is not summarized
summary(wyoming_all_years, cat_vars = "JWMNP")  # JWMNP is numeric
summary(wyoming_all_years, numeric_vars = "WAGP")  # not in the data
summary(wyoming_2022, numeric_vars = "JWAP")  # times are left out
summary.census(select(wyoming_all_years, -PWGTP))  # no weights to use
```


## Plotting a census tibble
 
The plot method draws a boxplot of one numeric variable at each level of one categorical variable. The user has to name both, as column-name strings. The categorical variable can be any factor in the tibble, including `year`. The numeric variable can be any numeric column other than `PWGTP`, and the time variables are left out.
 
The plot uses `weight = PWGTP` inside `aes()`, so each record counts as the number of people it represents and the boxes show the weighted quartiles. People with a missing value for the numeric variable are left out of the plot. `.data[[cat_var]]` allows the method to use column names supplied as strings. 
 
Find which columns can be plotted and check that the user named exactly one of each kind.
 
```{r}
#| label: plot-helpers
 
# the columns a census tibble can be plotted by: factors for the categories,
# and numeric columns other than the weight (times are left out) for the values
census_variables <- function(data) {
  is_number <- vapply(data, function(col) is.numeric(col) && !is.period(col),
                      logical(1))
  is_factor <- vapply(data, is.factor, logical(1))
  list(numeric     = setdiff(names(data)[is_number], "PWGTP"),
       categorical = names(data)[is_factor])
}
 
# like check_choices(), but the user has to give exactly one variable
check_one_choice <- function(value, allowed, arg_name) {
  if (length(value) != 1) {
    stop(arg_name, " must be a single variable name. Choose from: ",
         paste(allowed, collapse = ", "), ".", call. = FALSE)
  }
  check_choices(value, allowed, arg_name)
}
```
 
The plot method checks the inputs, then builds the plot.
 
```{r}
#| label: plot-census
 
# plot method for census tibbles: a weighted boxplot of num_var at each level of cat_var.
plot.census <- function(x, cat_var, num_var) {
  if (missing(cat_var) || missing(num_var)) {
    stop("give one categorical variable (cat_var) and one numeric variable ",
         "(num_var).", call. = FALSE)
  }
  if (!"PWGTP" %in% names(x)) {
    stop("the data needs a PWGTP column to use as weights.", call. = FALSE)
  }
  options <- census_variables(x)
  check_one_choice(cat_var, options$categorical, "cat_var")
  check_one_choice(num_var, options$numeric, "num_var")
  if (!requireNamespace("quantreg", quietly = TRUE)) {
    stop("weighted boxplots need the quantreg package. Install it with ",
         "install.packages(\"quantreg\").", call. = FALSE)
  }
 
  ggplot(x, aes(x = .data[[cat_var]], y = .data[[num_var]], weight = PWGTP)) +
    geom_boxplot(na.rm = TRUE) +
    labs(x = cat_var, y = num_var,
         title = paste("Weighted", num_var, "by", cat_var),
         subtitle = "Each record is weighted by PWGTP")
}
```
 
Calling `plot()` on the multi-year Wyoming data with `year` as the categories shows how travel time changed across the surveys.
 
```{r}
#| label: example-plot-year
 
plot(wyoming_all_years, cat_var = "year", num_var = "JWMNP")
```
 
The variables can also be given by position. This plot splits the same travel times by sex.
 
```{r}
#| label: example-plot-sex
 
plot(wyoming_all_years, "SEX", "JWMNP")
```
 
A different numeric variable and data set work the same way. This plot shows age by sex for North Carolina in 2024.
 
```{r}
#| label: example-plot-age
 
plot(nc_2024, cat_var = "SEX", num_var = "AGEP")
```
 
### Calls that should fail
 
The following examples test the main validation errors.
 
```{r}
#| label: example-plot-errors
#| error: true
 
plot(wyoming_all_years)  # both variables are required
plot(wyoming_all_years, cat_var = "year")  # num_var is missing
plot(wyoming_all_years, cat_var = "JWMNP", num_var = "JWMNP")  # JWMNP is numeric
plot(wyoming_all_years, cat_var = "year", num_var = "SEX")  # SEX is a factor
plot(wyoming_all_years, cat_var = c("year", "SEX"), num_var = "JWMNP")  # only one categorical
plot(wyoming_all_years, cat_var = "year", num_var = "PWGTP")  # the weight is not plotted
plot(wyoming_2022, cat_var = "SEX", num_var = "JWAP")  # times are not plotted
```



# Investigation: Commute Time, Age, and Education in North Carolina
 
This section uses the functions built above to ask one question about North Carolina: **do commute time and age differ by education level, and did that change from 2021 to 2024?**
 
Commute time (`JWMNP`) is missing for anyone who is not a worker and for workers who work from home, so every commute average here describes people who travel to work.

Education (`SCHL`) is the highest level a person has reached, and it has 25 levels. Children are on the scale too, in grades up to 12th, and people under 3 years old have their own "N/A" level.

Age and education are linked by construction. A person cannot have a college degree at age 10, so some of the age differences between education levels only reflect who has had time to finish school.
 
## Getting the data
 
One call to `get_pums_years()` downloads all four years for North Carolina, the default geography, with commute time and age as the numeric variables and education as the categorical variable. The result is saved to `data/nc_2021_2024.rds`.
 
```{r}
#| label: investigation-data
 
if (!file.exists("data/nc_2021_2024.rds")) {
  nc_2021_2024 <- get_pums_years(years = 2021:2024,
                                 numeric_vars = c("JWMNP", "AGEP"),
                                 cat_vars = "SCHL")
  saveRDS(nc_2021_2024, "data/nc_2021_2024.rds")
}
 
nc_2021_2024 <- readRDS("data/nc_2021_2024.rds")
nc_2021_2024
```
 
## Summaries
 
`summary()` with no other arguments reports the weighted mean and standard deviation of commute time and age, and the estimated number of people and the number of sample records at each level of education, year, and the geography column.
 
```{r}
#| label: investigation-summary
 
summary(nc_2021_2024)
```
 
Across the four years, the average commuter in North Carolina travels about 25.2 minutes to work, with a standard deviation of 21.1 minutes, so commute times vary widely. The average age is 39.7 years. The weighted counts by year are 10.55 million (2021), 10.70 million (2022), 10.84 million (2023), and 11.05 million (2024). These numbers are similar to that of North Carolina's population, suggesting the weights are working as intended. The counts by education and by state add all four years together, totalling 43.1 million.
 
The summary reports one overall mean, so to compare years the data is filtered to one year at a time. `filter()` keeps the `census` class, so `summary()` dispatches to the census method. This loop summarizes each year and puts the weighted means side by side.
 
```{r}
#| label: investigation-means-by-year
 
means_by_year <- sapply(levels(nc_2021_2024$year), function(y) {
  summary(filter(nc_2021_2024, year == y), cat_vars = "year")$means
})
means_by_year
```
 
Average commute time rose every year, from 24.6 minutes in 2021 to 25.1, 25.3, and 25.8 minutes in 2024. That is an increase of 1.2 minutes, or about 5 percent. Average age rose more slowly, from 39.5 to 39.8 years, an increase of about 0.3 years. This data shows that the commute times increased but not why they increased.
 
## Plots
 
The plot method draws a weighted boxplot. Commute time by survey year shows whether the overall distribution moved.
 
```{r}
#| label: investigation-plot-year
 
plot(nc_2021_2024, cat_var = "year", num_var = "JWMNP")
```
 
The boxes are almost identical across the four years. The median commute is 20 minutes in every year and the upper quartile is 30 minutes in every year. The one change is in 2024, where the lower quartile rises from about 12 to about 14 minutes, so very short commutes became less common. The mean rose from 24.6 to 25.8 minutes while the median stayed at 20, so the increase comes from changes away from the middle of the distribution. The dots above the whiskers are drawn one per record without weights, so they show which commute times occur, not how common they are.
 
Education has 25 levels, so the plots below are turned sideways with `coord_flip()`. Returning a `ggplot` object.
 
```{r}
#| label: investigation-plot-commute-education
#| fig-height: 8
 
plot(nc_2021_2024, cat_var = "SCHL", num_var = "JWMNP") + coord_flip()
```
 
Among people with 12th grade or more, the commute distributions look alike. The median is 20 minutes at every level from 12th grade with no diploma through the doctorate, and the middle half of commutes runs from about 10 to 15 minutes up to 30 minutes. The lower quartile is a little lower, about 10 minutes, for people with a professional degree and for people with some college credit but no degree. The grade-school levels, from nursery school through 11th grade, are based on a handful of adult workers as children do not commute to work. Those boxes should not be interpreted. The N/A level for children under 3 has no commute data and shows an empty row.
 
The same data with age as the numeric variable shows the link between age and education.
 
```{r}
#| label: investigation-plot-age-education
#| fig-height: 8
 
plot(nc_2021_2024, cat_var = "SCHL", num_var = "AGEP") + coord_flip()
```
 
The age distribution follows the education categories partly due to `SCHL` includes children who have not yet finished their education. The grade-level categories contain both children and a similar number of adults whose education stopped at that level. "No schooling completed" has a wide box, about 3 to 37 years with a median near 5, because it includes children who have not started school as well as adults. From 12th grade with no diploma up through the doctorate, the medians sit in a narrow band of about 43 to 51 years. The highest degrees are somewhat older. The median is about 49 to 51 for master's, professional, and doctorate degrees, compared with about 46 for a bachelor's degree.
 
## Did it change from 2021 to 2024?
 
This table computes the weighted mean commute time and age at each education level in 2021 and in 2024, using the same weighted_mean() as the summary method, and the change between them. It covers adults 25 and older because childrens grade levels have almost no commuters. The last two columns give the number of sample records with a commute time. Levels with small counts can be read with caution.
 
```{r}
#| label: investigation-change-table
 
education_change <- nc_2021_2024 |>
  filter(year %in% c(2021, 2024), AGEP >= 25) |>
  group_by(SCHL, year) |>
  summarize(commute   = weighted_mean(JWMNP, PWGTP),
            age       = weighted_mean(AGEP, PWGTP),
            commuters = sum(!is.na(JWMNP)),
            .groups = "drop") |>
  pivot_wider(names_from = year, values_from = c(commute, age, commuters)) |>
  mutate(commute_change = commute_2024 - commute_2021,
         age_change     = age_2024 - age_2021) |>
  select(SCHL, commute_2021, commute_2024, commute_change,
         age_2021, age_2024, age_change, commuters_2021, commuters_2024)

knitr::kable(education_change, digits = 1)
```
 
Education separates commute times by only a few minutes. Among the levels with at least 500 commuting records in both years, the average commute ranges from 22.8 to 26.9 minutes in 2021 and from 24.4 to 27.6 minutes in 2024. People with a professional degree have the shortest commute in both years, at 22.8 and 24.4 minutes. Every one of these levels had a longer average commute in 2024 than in 2021. The largest increases were for people with some college but less than a year (+2.3 minutes), a master's degree (+1.9), a bachelor's degree and a professional degree (+1.6 each), and a regular high school diploma (+1.5). A GED, an associate's degree, and 1 or more years of college credit without a degree each changed by 0.3 minutes or less. Average age among adults at these levels changed by between −0.2 and +0.9 years. The commute averages cover adults who travel to work, while the age averages cover all adults 25 and older at that level.

The levels below a high school diploma give noisy averages. In the grade-school levels there are between 2 and 58 commuting records in a year, so large swings, such as +10.6 minutes for nursery school (4 and 12 records) or −9.0 minutes for Grade 4 (33 and 23 records), reflect a handful of people and should be ignored.
 
## Conclusion
 
Among adults in North Carolina, commute time differs only slightly by education level. At the levels with many commuters the average commute is between about 23 and 27 minutes, the median is 20 minutes at every level from 12th grade through a doctorate, and people with a professional degree have the shortest commutes. Age differs by education mostly because of how education is measured, since children appear on the scale up to 12th grade. Among adults, average age only varies from about 49 to 54 years across the levels, with bachelor's degree holders the youngest.

Between 2021 and 2024 the average commute rose 1.2 minutes, from 24.6 to 25.8, while the median stayed at 20 minutes and the lower quartile rose from about 12 to about 14 minutes. Every education level with at least 500 commuting records in both years had a longer average commute in 2024, by 0.1 to 2.3 minutes. The largest increases were for people with some college, a master's degree, a bachelor's degree, or a professional degree. Average age changed minimally. The data cannot show why commutes got longer, and the biggest changes in the table come from education levels with very few commuters.
